White Wine Quality Exploration by Ying Xiong
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(grid)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("~/xyCode/udacity-data-analyst/explore-and-summarize")
df <- read.csv("data/wineQualityWhites.csv")
```

# Univariate Plots and Analysis

### What is the structure of your dataset?
Below prints a simple summary of the data.

There are 4898 white wine observations in the dataset with 13 variables in total, including an index variable (named "X"), the "quality" variable, and 11 other variables describing the chemical properties of the wine.

The `quality` of the wine is an integer variable which has has a min 3.0 and max 9.0, with a median 6.0 and mean 5.878.

All the chemical property variables are floating numbers. They are of different unit and therefore lie in widely different range. For example, the `chlorides` variable has a small range from 0.009 to 0.346, while the `total.sulfur.dioxide` variable has a large range from 8.0 to 440.0.

```{r echo=FALSE, Univariate_Structure}
str(df)
summary(df)
```

### What is/are the main feature(s) of interest in your dataset?

The main features in the data set are alcohol and quality. I suspect alcohol and some combination of other variables can be used to build a predictive model to the wine quality.

Below we plot the histogram for the `quality` variable. The variable is discrete, but we can see its histogram has a typical normal distribution shape.

```{r echo=FALSE, Univariate_Quality}
qplot(x = quality, data=df) +
  scale_x_discrete(limits = c(3,9), breaks = seq(3,9,1))
```

Similarly, we plot the histogram for the `alcohol` variable. The distribution is plotted at different `binwidth`, so that we can look at data with different "resolution". At the coarse level (`binwidth=1`), we see that it follows a skewed distribution with most number of samples in [9, 10], followed by [10, 11], and then [11, 12], etc. At the fine level (`binwidth=0.1`), we see more irragularities of the distribution with multiple spikes, say at [9.0, 9.1], [9.5, 9.6], [10.0, 10.1], etc.

```{r echo=FALSE, Univariate_Alcohol}
p1 <- qplot(x = alcohol, data=df, binwidth=0.1) +
  scale_x_continuous(limits = c(8,15), breaks = seq(8,15,1))
p2 <- qplot(x = alcohol, data=df, binwidth=0.2) +
  scale_x_continuous(limits = c(8,15), breaks = seq(8,15,1))
p3 <- qplot(x = alcohol, data=df, binwidth=0.5) +
  scale_x_continuous(limits = c(8,15), breaks = seq(8,15,1))
p4 <- qplot(x = alcohol, data=df, binwidth=1) +
  scale_x_continuous(limits = c(8,15), breaks = seq(8,15,1))

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Features such as `residual.sugar`, `sulphates`, `pH`, `chlorides` will likely contribute to the wine quality and will support our investigation.

### Did you create any new variables from existing variables in the dataset?

Currently I found it not necessary to create any new variables from existing ones in the dataset.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

During the investigation, I found the distribution of `chlorides` variable has an unusual distribution. From the histogram shown below, we see that the majority of samples lie in the range of [0, 0.1] in a normal distribution shape, but there are a small number of outliers that lie far beyond this normal range (up to 0.34), which indicates this is a long-tail distribution.

In order to better visualize this distribution, we tried two approaches
1. Cut off the samples that are beyond 0.1, and only "zoom in" to look at those in the "regular range";
2. Plot the distribution in a `log10` scale.

```{r Univariate_Chlorides}
p1 <- qplot(x = chlorides, data=df, binwidth=0.005,
            main = "Histogram of all samples in linear scale.")

p2 <- qplot(x = chlorides, data=df, binwidth=0.005,
            main = "Histogram of samples with value smaller than 0.1.") +
  scale_x_continuous(limits = c(0, 0.1))

p3 <- ggplot(aes(x = chlorides), data=df) + geom_histogram(binwidth = 0.05) +
  scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5)) +
  ggtitle("Histogram of all samples plotted in log10 scale.")

grid.arrange(p1, p2, p3, ncol = 1)
```


# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
